DPhil Thesis Proposal: Semantics of Probabilistic Functional Languages

### 2021-03-18 thesis proposal/thesis proposal.tex
Add introduction and start prior work section
I know I need to add references. I'm planning to do that after most
of the writing.

\documentclass[titlepage]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\newcommand{\tY}{\mathsf{Y}}
\newcommand{\tsample}{\mathsf{sample}}

@@

\begin{document}
\maketitle
\section{Introduction}

%Existing approaches:
    %brief summary, axes of difference: operational, denotational, distribution-based, trace-based
    %Operational Semantics
    %Denotational Semantics
        %Denotational Semantics in the non-probabilistic case, types and inconsistencies
        %QBSes
    %Distribution-based Semantics
    %Trace-based Semantics

>>>>

\documentclass[titlepage]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\newcommand{\tY}{\mathsf{Y}}
\newcommand{\tsample}{\mathsf{sample}}
\newcommand{\tscore}{\mathsf{score}}

@@

\begin{document}
\maketitle
\section{Introduction}
The augumentation of lambda calculus with random sampling and conditioning, the fundamental constructs of probabilistic programming, can be interpreted formally in many different ways, and provides certain unique challenges in defining the language's semantics. The standard approach to defining random things, where all the random values are considered as functions from a probability space, has the problem that the random samples, which should form the basis of this probability space, are not specified in advance. There are various possible schemes of organising the probability space such that this data is available, known as trace or entropy space semantics. The conditioning construct, which modifies the posterior probability of an event, fits somewhat unnaturally in this framework, and has to be specified as an extra step that modifies the probabilities at the end.

Alternatively, the distribution of results can be specified more directly as a distribution. This omits any information on joint distributions from the semantics, but this itself can be in some ways an advantage, as a subterm can be duplicated without being correlated to itself. It would be natural to attempt to use the category of measurable spaces, however, that turns out to not be cartesian closed. This motivates the introduction of quasi-Borel spaces as an alternative way in which a distribution can be defined, which has a more restrictive set of morphisms and does not have this issue, but is nevertheless still able to represent all the functions definable as programs.

For my thesis, I plan to extend the existing literature and my own previous work in defining the semantics of probabilistic functional languages.

\section{Existing approaches}
The various existing approaches to defining the semantics of probabilistic functional languages can generally be categorised as operational or denotational (depending on whether it works by defining the state of the program at each step, or by defining the meanings of subprograms independently and combining them together) (this distinction applies to functional languages more generally, rather than being specific to probabilistic languages), and distribution-based or trace-based (depending on whether only the distribution of outputs is defined, or the output that occurs is defined as a function of a random input).

As a running example, I will be using a simply-typed call by value language with a random sampling construct in the form of a probabilistic choice $M \oplus N$, a conditioning construct $\tscore(r)$ for non-negative real constants $r$, and countably many primitive values of some base type $\mathsf T$. Although this language is rather deficient (there is no way to use continuous distributions, for example, and no recursion), it will be sufficient to illustrate the different styles of semantics.

\subsection{Operational Semantics}
In operational semantics, the meaning of the program is defined by repeatedly transforming a term into other terms, until it reaches a value, a term which is considered finished, which is the output. A small-step trace-based operational semantics of the example language can be defined as follows.

First, an \emph{environment} is defined as
\begin{equation*}
E := \cdot \mid E M \mid V E,
\end{equation*}
so that an environment is a term with a single hole, where $M$ is a term and $V$ is a value. Substituting terms into environments is defined in the usual way. The single-step reduction relation can then be defined as
\begin{align*}
(E[(\lambda x. M) V], t, s) & \to (M[V/x], t, s) \\
(E[M \oplus N], t, s) & \to (E[M], 0:t, s) \\
(E[M \oplus N], t, s) & \to (E[N], 1:t, s) \\
(E[\tscore(r)], t, s) & \to (E[(\lambda x. x)], t, rs),
\end{align*}
where $t$ is a list of booleans, $:$ is the cons operator, and $s$ is a real number. The probability of $M$ reaching a value $V$ can then be defined as
\[
\sum_{(M,[],1) \to^* (V,t,s)} 2^{-\mathrm{length}(t)} s.
\]

Alternatively, this could be defined in a different but closely related version of operational semantics, the big-step semantics:
\begin{align*}
\frac{(M, t_0, s_0) \Downarrow (\lambda x. M', t_1, s_1) \; (N, t_1, s_1) \Downarrow (N', t_2, s_2) \; (M'[N'/x], t_2, s_2) \Downarrow (O, t_3, s_3)}{(M N, t_0, s_0) \Downarrow (O, t_3, s_3)} \quad
\frac{(M, 0:t_0, s_0) \Downarrow (M', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (M', t_1, s_1)} \quad
\frac{(N, 1:t_0, s_0) \Downarrow (N', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (N', t_1, s_1)} \quad
\frac{}{(\tscore(r), t, s) \Downarrow (\lambda x. x, t, rs)}.
\end{align*}
The notation in this case could be simplified somewhat, but that would obscure the connection to the small-step semantics. If $V$ is a value, then $(M, t_0, s_0) \to^* (V, t_1, s_1)$ iff $(M, t_0, s_0) \Downarrow (V, t_1, s_1)$.

\subsection{Denotational Semantics}
        %Denotational Semantics in the non-probabilistic case, types and inconsistencies
        %QBSes
    %Distribution-based Semantics
    %Trace-based Semantics
